A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-05-06 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Final edits after discussion with co-authors, then resubmission to Geographical Analysis.

From a small outbreak linked to a live animal market at the end of 2019
to a global pandemic in a matter of weeks, the SARS-CoV2 virus (and
COVID-19, the disease it causes), has threatened to overrun health
systems around the world. In efforts to contain the spread of the
disease, numerous governments in many regions and nations have either
recommended or mandated social distancing measures, and as of this
writing, millions of people in five continents shelter in place. There
are encouraging signs that these measures have mitigated the spread of
the virus (e.g., Lancastle, 2020; Lewnard and Lo, 2020; Wilder-Smith and
Freedman, 2020). Even so, this has come at a high cost, and the
consequences for all spheres of economic, social, and cultural life have
been dire (e.g., Fernandes, 2020; Luo and Tsang, 2020). As a result,
there is a sense of urgency to anticipate the progression of the
pandemic, in order to plan for progressive lifting of restrictions to
movement and social contact (e.g., Kissler et al., 2020). Needless to
say, this has become a delicate, and politically charged, balancing act
between public health and the economy (Gong et al., 2020).

A salient question in the debate on how and when to ease social
distancing measures is whether the virus will display seasonal
variations. Existing research on similar pathogens suggests that the
virus could be more stable and potentially easier to transmit in
conditions of low temperature and low humidity. While encouraging this
is encouraging, it is important to keep in mind that ``not all seasonal
respiratory viruses experience the same spatiotemporal patterns'' (Ángel
Solá et al., 2020, sec. 4). This urges caution when extrapolating from
known viruses. The evidence in this respect is as yet inconclusive, and
although easing restrictions as the weather warms may appear tempting,
doing so prematurely could well undo weeks or possibly months of costly
measures.

It is not surprising, given the stakes involved, that this issue has
already triggered a lively debate. The current state of knowledge is was
well-summarized by the National Academy of Sciences, Engineering, and
Medicine in the U.S. (see National Academies of Sciences, Engineering
and Medicine, 2020). Engaged by the Office of the Executive for guidance
on this matter, this organization concludes that: ``{[}some{]} limited
data support a potential waning of cases in warmer and more humid
seasons, yet none are without major limitations\ldots Additional studies
as the SARS-CoV2 pandemic unfolds could shed more light on the effects
of climate on transmission'' (p.~6). To further complicate matters, much
of the relevant work has yet to be peer-reviewed (see for instance the
challenge of Harbert et al. (2020) to Araujo and Naimi (2020)).

With the above considerations in mind, our objective with this paper is
to investigate the influence of environmental factors, concretely
temperature, humidity, and sunshine, on the progression of the pandemic.
We adopt a population health approach, and report results from a
spatio-temporal model of the incidence of COVID-19 in fifty provinces in
Spain, one of the countries hardest hit by the pandemic. We combine data
on reported cases of the disease with metereological information, to
create a spatio-temporal dataset covering a period of 30 days. We then
join this dataset with provincial-level economic and demographic
information to act as controls to our key environmental variables. These
data are analyzed using a spatial Seemingly Unrelated Regressions (SUR)
approach, which allows us to model incidence of COVID-19 as a contagion
process.

@@

Please note that this paper is prepared as a reproducible research
document. The source \texttt{R} markdown document, as well as all data
and code needed to reproduce/review/extend the analysis can be obtained
from a public
repository\footnote{https://github.com/paezha/covid19-environmental-correlates}.

@@

The global emergence of infectious diseases is mostly driven by
environmental, ecological, and socio-economic factors (Jones et al.,
2008). In the case of SARS-CoV2, the ecological factors include the
interaction between humans and wildlife. Once transmission of a disease
begins to happen between humans, socio-economic and environmental
factors become increasingly important. As noted in the introduction, the
focus of the paper is on environmental variables, namely temperature,
humidity, and sunshine.

Much of what is known about the potential seasonality of SARS-CoV2 is a
result of research on other pathogens. Earlier, diverse studies have
shown the effect of temperature and humidity on the incidence of
influenza (e.g., Mäkainen et al., 2009; Jaakkola et al., 2014; Kudo et
al., 2019). Jaakkola et al.~(2014), for example, found that a decrease
of temperature and absolute humidity precedes the onset of symptoms of
influenza A and B viruses by 3 days in places where the temperature is
low. After the 2002-2004 outbreak of SARS, researchers also began to
investigate the relationship between these factors and SARS-CoV2
(Casanova et al., 2010; Chan et al., 2011). Casanova et al.~(2010), for
instance, used two surrogates (i.e., the gastroenteritis TGEV and mouse
hepatitis MHV viruses), to find that virus inactivation was more rapid
at temperatures of 20°C than 4°C, and at 40°C than 20°C; in terms of
humidity, these researchers reported that survival of the virus was
lower at moderate relative humidity levels. Chan et al.~(2011) also
found that viability of the virus that causes SARS is also lost at
higher temperatures (\textgreater38°C) and relative humidity superior to
95\%.

Whether these results from laboratory experiments will hold in general
remains uncertain. At a global scale, de Ángel Solá et al.~(2020) see
less risk in the Caribean Basin; however, Coelho et al.~(Coelho et al.,
2020) warn that at least during the exponential phase, expansion of the
virus is not driven by climate. Similarly, whereas Araujo and Naimi
(2020) argue that spread of SARS-CoV2 will likely be constrained by
climate, Harbert et al.~(2020) remain unconvinced that spatial modelling
can currently discriminate the distribution of the disease on the basis
of climate, at least in the United States. Yao et al.~(2020), examined
data from China and came to the conclusion that neither temperature nor
ultraviolet indices had an association with transmission of COVID-19.
This is despite previous research that has linked less exposure to UVB
radiation to higher prevalence and severity of acute respiratory tract
infections (Zittermann et al.~2016; Dąbrowska-Leonik et al.~2018; Dinlen
et al.~2016; Mathyssen et al.~2017; Esposito and Lelii 2015; Jat 2017;
Moriyama, Hugentobler, and Iwasaki 2020).

In addition to the environmental variables above, we also consider a set
of potential socio-economic and demographic confounders.

The first variable that we consider is GDP per capita. Much has been
said about globalization and the spread of infectious
disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p. A18)}.
The growth in global connections has presented a challenge to spatial
approaches in the initial stages of disease management, when the cause
of a disease may still be unclear, but the plane has already departed
(Zhou and Coleman, 2016). In reference to the earlier outbreak of SARS,
van Wagner (Van Wagner, 2008) chronicles how Toronto's status as a
global city turned out to be a vulnerability in this respect. In our
case, we think of GDP per capita as a marker of a region's relative
position in a network of global cities, and its potential to be further
ahead in the trajectory of the pandemic. Furthermore, wealthier regions
also tend to concentrate more activities that produce non-traded goods,
including building and construction (Hallet, 2002). Therefore, it is
possible that wealthier regions remain relatively more active even
during a lockdown. On the other hand, it is also possible that less
wealthy regions have a higher proportion of workers in manual
occupations that cannot telework, and therefore have more difficulties
complying with shelter-in-place orders.

Secondly, we consider percentage of older adults (over 65) in a region.
Early evidence regarding COVID-19 suggests that the case rate mortality
is higher at older ages (e.g.~The Novel Coronavirus Pneumonia Emergency
Response Epidemiology Team, 2020). However, it is not clear that a
relatively large population of older adults necessarily translates into
higher transmission of the infection. The tool of choice in containing
the spread of the disease has been social distancing. In this respect,
the evidence from the field of transportation is that older adults tend
to travel less frequently, for shorter distances, and have higher rates
of immobility (e.g., Roorda et al., 2010; Morency et al., 2011; Sikder
and Pinjari, 2012). In other words, many older adults are, whether by
preference or otherwise, already in a form of social isolation. Social
distancing during the pandemic may actually reinforce that condition for
them, as suggested by the analysis of age-structured social contact in
India, China, and Italy of Sing and Adhikari (2020). Accordingly, our
expectation is that higher percentages of older adults will tend to have
similarly low levels of social contact (the age-structured matrix of
social contact in Spain is similar to Italy; see Prem et al., 2017).

@@

The last variable that we consider as a control is the presence of mass
transit systems in a province. Every province in Spain offers some form
of public transportation, however only five provinces have higher order
systems of mass mobility (e.g.~metro or subway), namely Barcelona,
Madrid, Sevilla, Valencia, and Bizkaia. Public transportation has been
hypothesized to relate to the spread of contagious disease by some
researchers using agent-based approaches and simulation (e.g., Perez and
Dragicevic, 2009; Wang et al., 2010), and while we find scant evidence
of a link in the literature, the idea is intuitively appealing. After
all, unlike the isolation that a car offers to travellers, most mass
transit system are cauldrons of social contact.

@@

The first reported case of COVID-19 in Spain was on January 31st, 2020,
when a German tourist in the Canary Islands tested positive for the
virus. However, it was still a few weeks before the first domestic case
was reported, on February 27th in Sevilla province (Andalusia). In a
short period of time, as testing started to ramp up, it became clear
that an outbreak was flaring. By March 11th the World Health
Organization (WHO) declared COVID-19 officially a pandemic. This
declaration marked a turning point for the public in Spain too. As of
March 13th, the number of cases of COVID-19 reported in Spain was 4,473,
with a majority of cases (1,990) concentrated in Madrid: these numbers
were at the time the worst outbreak in Europe after Italy. In response
to the situation, on March 13th the Spanish National Government declared
a state of emergency, to go into effect on Saturday March 14th. As part
of the state of emergency restrictions to most activities were imposed,
with the exception of essential services (e.g.~food, health) and some
economic subsectors of industry and construction. A few days later, on
March 17th, Spain closed its lands borders to allow entry only to
returnee nationals and permanent residents. The lockdown was further
hardened between March 30th and April 12th (including the Easter weekend
of April 10th-12th) and during this period only essential activities
were allowed. During this period, there was a dramatic reduction in
overall mobility, both within provinces as between
\footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.

@@

Our dataset includes information about the daily number of cases of
COVID-19 reported in Spain at the provincial level (NUTS3 in Eurostat
terminology) for the period between March 13th and April 11th,
inclusive. For our purposes, we consider positive cases reported, but
excluding symptomatic cases diagnosed by a doctor without a Polymerase
Chain Reaction (PCR) test. The Spanish National Government publishes
periodic updates at the regional level (NUTS2) and the information is
also released at the provincial level as part of a collaborative project
by geovoluntarios.com\footnote{https://www.geovoluntarios.org/},
ProvidencialData19\footnote{https://www.datoscovid.es/pages/providencialdata19},
and ESRI España. This information is compiled from several sources,
mainly the official web pages of the Spanish regional goverments, as
documented in Centro de Datos
Covid-19\footnote{https://www.datoscovid.es/pages/sobre-la-iniciativa}.
In addition, we consider two sets of explanatory variables. The first
one, and the focus of this research, is set of two environmental
variables, namely temperature and humidity, which are collected from
official sources (i.g., \emph{AEMET}, the state meteorology agency, and
\emph{MAPA}, the ministry of agriculture, fisheries, and food). The
second set provides some relevant controls for multivariate analysis,
and refers to economic and demographic attributes of the province (also
collected from official sources, i.e., INE, the national statistics
institute). Table \ref{tab:descriptive-statistics} shows the descriptive
statistics and the provenance of the data.

@@

An important aspect of working with environmental data such as
temperature and humidity is the incubation period of the disease. Lauer
et al.~(2020) report for the case of COVID-19 a median incubation period
of 5.7 days (with a confidence interval between 4.9 to 7.8 days). The
vast majority of cases (95\%) develop symptoms between 2.6 days (CI, 2.1
to 3.7 days) and 12.5 days (CI, 8.2 to 17.7 days). For this reason, we
judge it best to use lagged values of the environmental variables. We
test different time lags as follows. We consider a lagged 8-day average,
from date-minus-12 to date-minus-5 days (hereafter \emph{lag8}).
Secondly, we consider a lagged 11-day average, from date-minus-12 to
date-minus-2 days (hereafter \emph{lag11}). Finally, to account for the
likely duration of incubation, we consider a lagged 11-day
\emph{weighted} average, from date-minus-12 to date-minus-2 days
(hereafter \emph{lag11w}). The weights for this variable are calculated
using the parameters of the log-normal distribution reported by Lauer et
al.~(2020), i.e., a log-mean of 1.621 and a log-standard deviation of
0.418. With these weights, the environmental variables at date-minus-2
and date-minus-12 days are weighted as 0.041 and 0.009 respectively,
whereas the environmental variables at date-minus-5 days are weighted as
0.194.

@@

Figure \ref{fig:weekly-average-incidence-map} shows the geographical
variation in the incidence of COVID-19 in Spain, as well as the temporal
progression of the disease in weekly averages. Our analysis begins on
March 13, which is the first date when every province had reported at
least one case of COVID-19. It can be seen that the highest incidence at
this early date was in the provice of Álava, in the North of Spain.
While not the most populous province, with a population of only 331,549,
Álava has the highest GDP per capita of all provinces. Vitoria, its main
city, is the capital of the Basque Country and has been the focus of
efforts, along with San Sebastian and Bilbao, to develop a ``Global
Basque City'' (Meijers et al., 2008). The other early focus of the
pandemic in Spain was Madrid, which is the most populous province in the
country and has the second highest GDP per capita after Álava. The early
outbreaks in these two provinces can be traced throughout the
progression of the pandemic over time, although by the end of the period
under consideration, other provinces had matched and even surpased their
incidence rates, including Segovia and Soria to the north of Madrid, and
Ciudad Real and Albacete to the south.

@@

Figure \ref{fig:descriptives-temperature} shows the distribution of the
environmental variables in Spain. For ease of visualization we aggregate
the provinces by Autonomous Community. Each box-and-whisker in the
figure represents the distribution of the variable for a community over
the 30-day period. In the plot, the communities have been sorted by
latitude, so that Principado de Asturias is the northernmost community,
and Canarias the southernmost. As seen in the figure, there is a
relatively wide range of values both within and between provinces over
the 30-day period examined. The top panel of the figure shows the
distribution of mean temperatures. The lowest mean temperature for a
community on any given day was 2.8°C, and the highest 22.4°C, for a
range of approximately 20 degrees. Likewise, there is a great deal of
variability in humidity, as seen in the middle panel of the figure,
where the lowest mean humidity for any community is 48.3 and the highest
is 99.6. Finally, the bottom panel displays mean daily hours of sunshine
in the community. This variable displays the most variability within
communities over time, but remains relatively constant across
communities. It is important to note that these are summaries by
community, and the actual values of these variables for the provinces
display somewhat more variability.

@@

Figure \ref{fig:daily-correlations} shows the distribution of daily
simple correlations of incidence of COVID-19 with the independent
variables (with the exception of Transit, which is a categorical
variable). These correlations are calculated after log-transforming all
variables. As previously discussed, the environmental variables have a
temporal lag and were calculated different time windows.

It can be seen in the figure that temperature (in its three forms) has
the highest simple correlation with incidence of COVID-19. After
temperature, GDP per capita has the highest positive correlation with
the dependent variable. The distribution of these correlations is also
quite tight over the 30-day period of the study. Hours of sunshine tends
to have a moderately high correlation with incidence of COVID-19,
however the distribution of these correlations is more spread, and in
some cases strays into negative values. A similar thing happens with
humidity, which also tends to display more day to day variation in the
correlation with the dependent variable. The percentage of older adults
shows a relatively tight distribution of day-to-day correlations, and is
negative. Population density, in contrast, tends to be negative, but is
relatively spread, and on some days, the simple correlation between
density and incidence of COVID-19 is weakly positive. Overall for the
period under examination, the pairwise correlations between these
variables and incidence are significant at \(p<0.05\), with the
exception of the three \(Sunshine_hours\) variables.

@@

To implement the SUR approach, we must define a matrix of spatial
weights \(W\). In this case, we consider neighborhoods based on
adjacency, based on the well-known queen criterion (two provinces are
adjacent if they share a boundary or touch at a vertex). We modify this
criterion in two cases. First, we make an allowance for adjacency
between the two islands in the Autonomous Community of Canarias in the
Pacific (Las Palmas and Santa Cruz de Tenerife), which we assume are
adjacent between each other. And secondly, we assume that Islas Baleares
in the Mediterranean are adjacent to three provinces in Pais Catalans
(i.e., Barcelona, Tarragona, and
Castello)\footnote{As a check for robustness, we also tested the rook criterion (two provinces are adjacent if they share a boundary, but not if they only touch at a vertex), and removed the three islands in the sample. The results (which can be consulted in the source R markdown document) are robust to the specification of $W$.}.

@@

After estimation, we compare the goodness of fit of the three SUR
models. Figure \ref{fig:goodness-of-fit} shows the equation-level
coefficient of determination \(R^2\), one for each time period/day. As
well, the overall coefficient of determination for the system is
reported for each model \(\text{pooled}-R^2\). The general trend is
identical for the three models, with the goodness-of-fit improving over
time and plateuing around a value of \(R^2\) of 0.6. Model 1
(\emph{lag8}) performs somewhat better in the first few equations/days,
when the goodness-of-fit is relatively poor, and then again in the last
few equations/days. Model 3 (\emph{lag11w}), in contrast, does not
perform well towards the end of the study period. The most balanced
model in terms of equation-level goodness-of-fit appears to be Model 2
(\emph{lag11}), and this impression is further supported by a slightly
higher value of the \(\text{pooled}-R^2\). The analysis using a lagged
moving average of the environmental variables is in line with the
incubation period reported by Lauer et al.~(2020), although the results
do not support the use of a weighted average. For the remainder of the
paper, we will adopt Model 2 (\emph{lag11}) as our best model. In the
following section we discuss the results of the analysis in more depth.

>>>>

From a small outbreak linked to a live animal market at the end of 2019
to a global pandemic in a matter of weeks, the SARS-CoV2 virus and
COVID-19, the disease it causes, have threatened to overrun health
systems around the world. In efforts to contain the spread of the
disease, numerous governments in many regions and nations have either
recommended or mandated social distancing measures, and as of this
writing, millions of people in five continents shelter in place. There
are encouraging signs that these measures have mitigated the spread of
the virus (e.g., Lancastle, 2020; Lewnard and Lo, 2020; Wilder-Smith and
Freedman, 2020). Even so, this has come at a high cost, and the
consequences for all spheres of economic, social, and cultural life have
been dire (e.g., Fernandes, 2020; Luo and Tsang, 2020). As a result,
there is a sense of urgency to anticipate the progression of the
pandemic, in order to plan for progressive lifting of restrictions to
movement and social contact (e.g., Kissler et al., 2020). Needless to
say, this has become a delicate, and politically charged, balancing act
between public health and the economy (Gong et al., 2020).

A salient question in the debate about how and when to ease social
distancing measures is whether the virus will display seasonal
variations. Existing research on similar pathogens suggests that the
virus could be more stable and potentially easier to transmit in
conditions of low temperature and low humidity. While this is
encouraging, it is important to keep in mind that ``not all seasonal
respiratory viruses experience the same spatiotemporal patterns'' (Ángel
Solá et al., 2020, sec. 4). This urges caution when extrapolating from
known viruses. The evidence in this respect is as yet inconclusive, and
although easing restrictions as the weather warms may appear tempting,
doing so prematurely could well undo weeks or possibly months of costly
measures.

It is not surprising, given the stakes involved, that this issue has
already triggered a lively debate. The current state of knowledge was
well-summarized by the National Academy of Sciences, Engineering, and
Medicine in the U.S. in a recent report (see National Academies of
Sciences, Engineering and Medicine, 2020). Engaged by the Office of the
Executive for guidance on this matter, this organization concluded that:
``{[}some{]} limited data support a potential waning of cases in warmer
and more humid seasons, yet none are without major
limitations\ldots Additional studies as the SARS-CoV2 pandemic unfolds
could shed more light on the effects of climate on transmission''
(p.~6). To further complicate matters, much of the relevant work has yet
to be peer-reviewed (see for instance the challenge of Harbert et al.,
2020; to Araujo and Naimi, 2020).

With the above considerations in mind, our objective with this paper is
to investigate the influence of environmental factors, concretely
temperature, humidity, and sunshine, on the progression of the pandemic.
We adopt a population health approach, and report results from a
spatio-temporal model of the incidence of COVID-19 in the coterminous
provinces in Spain, one of the countries hardest hit by the pandemic. We
combine data on reported cases of the disease with metereological
information, to create a spatio-temporal dataset covering a period of 30
days. We then join this dataset with provincial-level economic and
demographic information to act as controls to our key environmental
variables. These data are analyzed using a spatial Seemingly Unrelated
Regressions (SUR) approach, which allows us to model incidence of
COVID-19 as a contagion process.

@@

Please note that this paper is prepared as a reproducible research
document. The source \texttt{R} markdown document, as well as all data
and code needed to reproduce/review/extend the analysis can be obtained
from a public
repository\footnote{\url{https://drive.google.com/drive/u/1/folders/1d_40N_QXo2Fl14r3T3CHs84IED-B1rV_}}.

@@

The global emergence of infectious diseases is mostly driven by
environmental, ecological, and socio-economic factors (Jones et al.,
2008). In the case of SARS-CoV2, the ecological factors include the
interaction between humans and wildlife. Once transmission of a disease
begins to happen between humans, socio-economic and environmental
factors become increasingly important. As noted in the introduction, the
focus of the paper is on environmental variables, concretely three
related to meteorological conditions: temperature, humidity, and
sunshine.

Much of what is known about the potential seasonality of SARS-CoV2 is a
result of research on other pathogens. Earlier, diverse studies have
shown the effect of temperature and humidity on the incidence of
influenza (e.g., Mäkainen et al., 2009; Jaakkola et al., 2014; Kudo et
al., 2019). Jaakkola et al.~(2014), for example, found that a decrease
of temperature and absolute humidity precedes the onset of symptoms of
influenza A and B viruses by 3 days in places where the temperature is
low. After the 2002-2004 outbreak of SARS, researchers also began to
investigate the relationship between these factors and SARS-CoV
(Casanova et al., 2010; Chan et al., 2011). Casanova et al.~(2010), for
instance, used surrogates to find that virus inactivation was likely
more rapid at higher temperatures; in terms of humidity, these
researchers reported that survival of the virus was lower at moderate
relative humidity levels. Chan et al.~(2011) also found that viability
of the virus that causes SARS is also lost at higher temperatures
(\textgreater38°C) and relative humidity superior to 95\%.

Whether results from laboratory experiments will hold when the virus
circulates in the community remains uncertain. At a global scale, de
Ángel Solá et al.~(2020) see less risk from SARS-CoV2 in the Caribean
Basin; on the other hand, Coelho et al.~(Coelho et al., 2020) warn that
at least during the exponential phase, expansion of the virus is not
driven by climate. Similarly, whereas Araujo and Naimi (2020) argue that
spread of SARS-CoV2 will likely be constrained by climate, Harbert et
al.~(2020) remain unconvinced that spatial modelling can currently
discriminate the distribution of the disease on the basis of climate, at
least in the United States. Yao et al.~(2020), examined data from China
and came to the conclusion that neither temperature nor ultraviolet
indices had an association with transmission of COVID-19. This is
despite previous research that has linked less exposure to UVB radiation
to higher prevalence and severity of acute respiratory tract infections
(Zittermann et al.~2016; Dąbrowska-Leonik et al.~2018; Dinlen et
al.~2016; Mathyssen et al.~2017; Esposito and Lelii 2015; Jat 2017;
Moriyama, Hugentobler, and Iwasaki 2020).

In addition to the environmental variables above, from a population
health perspective it is also important to account for potential
socio-economic and demographic confounders.

To account for population-level factors, the first variable that we
consider is GDP per capita. Much has been written about globalization
and the spread of infectious
disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p. A18)}.
The growth in global connections has presented a challenge to spatial
approaches in the initial stages of disease management, when the cause
of a disease may still be unclear but the plane has already departed
(Zhou and Coleman, 2016). In reference to the earlier outbreak of SARS,
van Wagner (2008) chronicles how Toronto's status as a global city
turned out to be a vulnerability in this respect. In our case, we think
of GDP per capita as a marker of a region's relative position in a
network of global cities, and its potential to be further ahead in the
trajectory of the pandemic. Furthermore, wealthier regions also tend to
concentrate more activities that produce non-traded goods, including
building and construction (Hallet, 2002). Therefore, it is possible that
wealthier regions remain relatively more active even during a lockdown.
On the other hand, we cannot discount the possibility that less wealthy
regions have a higher proportion of workers in manual occupations who
cannot telework, and therefore have more difficulties complying with
shelter-in-place orders.

Secondly, we consider percentage of older adults (over 65) in a region.
Early evidence regarding COVID-19 suggests that the case rate mortality
is higher at older ages (e.g.~The Novel Coronavirus Pneumonia Emergency
Response Epidemiology Team, 2020). However, it is not clear that a
relatively large population of older adults necessarily translates into
higher transmission rates of the infection. The tool of choice in
containing the spread of the disease has been social distancing. In this
respect, the evidence from the field of transportation is that older
adults tend to travel less frequently, for shorter distances, and have
higher rates of immobility than most everyone, except the youngest
members of the public (e.g., Roorda et al., 2010; Morency et al., 2011;
Sikder and Pinjari, 2012). In other words, many older adults are,
whether by preference or otherwise, already in a form of social
isolation. Social distancing during the pandemic may actually reinforce
that condition for them, as suggested by the analysis of age-structured
social contact in India, China, and Italy of Sing and Adhikari (2020).
Since the age-structured matrix of social contact in Spain is similar to
Italy (see Prem et al., 2017), our expectation is that populations with
higher percentages of older adults will tend to have lower levels of
social contact and hence of incidence.

@@

The last variable that we consider as a control is the presence of mass
transit systems in a province. Every province in Spain offers some form
of public transportation, but only five provinces have higher order
systems of mass mobility (e.g.~metro or subway), namely Barcelona,
Madrid, Sevilla, Valencia, and Bizkaia. Public transportation has been
hypothesized to relate to the spread of contagious disease by some
researchers using agent-based approaches and simulation (e.g., Perez and
Dragicevic, 2009; Wang et al., 2010), and while we find scant evidence
of a link in the literature, the idea is intuitively appealing. After
all, unlike the isolation that a car offers to travellers, most mass
transit system are cauldrons of social contact.

@@

The first reported case of COVID-19 in Spain was on January 31st, 2020,
when a German tourist in the Canary Islands tested positive for the
virus. After this case, it was still a few weeks before the first
domestic case was reported, on February 27th in Sevilla province
(Andalusia). In a short period of time, as testing started to ramp up,
it became clear that an outbreak was flaring. By March 11th the World
Health Organization (WHO) declared COVID-19 officially a pandemic. This
declaration marked a turning point for the public in Spain too. As of
March 13th, the number of cases of COVID-19 reported in Spain was 4,473,
with a majority of cases (1,990) concentrated in Madrid: these numbers
were at the time the worst outbreak in Europe after Italy. In response
to the situation, on March 13th the Spanish National Government declared
a state of emergency, to go into effect on Saturday March 14th. As part
of the state of emergency restrictions to most activities were imposed,
with the exception of essential services (e.g.~food, health) and some
economic subsectors of industry and construction. A few days later, on
March 17th, Spain closed its lands borders to allow entry only to
returnee nationals and permanent residents. The lockdown was further
hardened between March 30th and April 12th (including the Easter weekend
of April 10th-12th) and during this period only essential activities
were allowed. During this period, there was a dramatic reduction in
overall mobility, both within provinces as between
\footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.

@@

Our dataset includes information about the daily number of cases of
COVID-19 reported in Spain at the provincial level (NUTS3 in Eurostat
terminology) for the period between March 13th and April 11th,
inclusive. For our purposes, we consider positive cases reported, but
exclude symptomatic cases diagnosed by a doctor without a Polymerase
Chain Reaction (PCR) test. The Spanish National Government publishes
periodic updates at the regional level (NUTS2) and the information is
also released at the provincial level as part of a collaborative project
by geovoluntarios.com\footnote{https://www.geovoluntarios.org/},
ProvidencialData19\footnote{https://www.datoscovid.es/pages/providencialdata19},
and ESRI España. This information is compiled from several sources,
mainly the official web pages of the Spanish regional goverments, as
documented in Centro de Datos
Covid-19\footnote{https://www.datoscovid.es/pages/sobre-la-iniciativa}.
We consider two sets of explanatory variables. The first one, and the
focus of this research, are the three environmental variables, collected
from official sources (i.g., \emph{AEMET}, the state meteorology agency,
and \emph{MAPA}, the ministry of agriculture, fisheries, and food). The
second set provides some relevant controls as discussed above, and are
also collected from official sources, i.e., INE, the national statistics
institute. Table \ref{tab:descriptive-statistics} shows the descriptive
statistics and the provenance of the data used in this research.

@@

An important aspect of working with environmental data such as
temperature, humidity, and hours of sunshine, is the incubation period
of the disease. Lauer et al.~(2020) report for the case of COVID-19 a
median incubation period of 5.7 days (with a confidence interval between
4.9 to 7.8 days). The vast majority of cases (95\%) develop symptoms
between 2.6 days (CI, 2.1 to 3.7 days) and 12.5 days (CI, 8.2 to 17.7
days). For this reason, we judge it best to use lagged values of the
environmental variables. We test different time lags as follows. We
consider a lagged 8-day average, from date-minus-12 to date-minus-5 days
(hereafter \emph{lag8}). Secondly, we consider a lagged 11-day average,
from date-minus-12 to date-minus-2 days (hereafter \emph{lag11}).
Finally, to account for the likely duration of incubation, we consider a
lagged 11-day \emph{weighted} average, from date-minus-12 to
date-minus-2 days (hereafter \emph{lag11w}). The weights for this
variable are calculated using the parameters of the log-normal
distribution reported by Lauer et al.~(2020), i.e., a log-mean of 1.621
and a log-standard deviation of 0.418. With these weights, the
environmental variables at date-minus-2 and date-minus-12 days are
weighted as 0.041 and 0.009 respectively, whereas the environmental
variables at date-minus-5 days are weighted as 0.194. These weights have
the effect of changing the contribution of daily values to the lagged
moving average. For instance, the temperature at date-minus-4-days is
weighted more heavily than the temperature at date-minus-10-days, as a
closer approximation of the conditions at the most likely time of
contagion before the disease became manifest.

@@

Figure \ref{fig:weekly-average-incidence-map} shows the geographical
variation in the incidence of COVID-19 in Spain, as well as the temporal
progression of the disease in weekly averages. Our analysis begins on
March 13. Albeit still low, the highest incidence at this early date was
in the provice of Álava, in the North of Spain. Álava is not the most
populous province, with a population of only 331,549, but it has the
highest GDP per capita of all provinces. Vitoria, its main city, is the
capital of the Basque Country and has been the focus of efforts, along
with San Sebastian and Bilbao, to develop a ``Global Basque City''
(Meijers et al., 2008). The other early focus of the pandemic in Spain
was Madrid, which is the most populous province in the country and has
the second highest GDP per capita after Álava. The early outbreaks in
these two provinces can be traced throughout the progression of the
pandemic over time, although by the end of the period under study, other
provinces had matched and even surpased their incidence rates, including
Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete
to the south.

@@

Figure \ref{fig:descriptives-temperature} shows the distribution of the
environmental variables in Spain. For ease of visualization we aggregate
the provinces by Autonomous Community. Each box-and-whisker in the
figure represents the distribution of the variable for a community over
the 30-day period. In the plot, the communities have been sorted by
latitude, so that Principado de Asturias is the northernmost community,
and Andalucia the southernmost. As seen in the figure, there is a
relatively wide range of values both within and between provinces over
the 30-day period examined. The top panel of the figure shows the
distribution of mean temperatures. The lowest mean temperature for a
community on any given day was approximately 3°C, and the highest about
20°C, for a range of approximately 17 degrees. Likewise, there is a
great deal of variability in humidity, as seen in the middle panel of
the figure, where the lowest mean humidity for any community is
approximately 48\% and the highest is close to 100\%. Finally, the
bottom panel displays mean daily hours of sunshine in the community.
This variable displays the most variability within communities over
time, but remains relatively constant across communities. It is
important to note that these are summaries by community, and the actual
values of these variables for the provinces display somewhat more
variability.

@@

Figure \ref{fig:daily-correlations} shows the distribution of daily
simple correlations of incidence of COVID-19 with the independent
variables (with the exception of Transit, which is a categorical
variable). These correlations are calculated after log-transforming all
variables. As previously discussed, the environmental variables have a
temporal lag and were calculated using different time windows.

It can be seen in the figure that temperature (in its three forms) has
the highest simple correlation with incidence of COVID-19. After
temperature, GDP per capita has the highest positive correlation with
the dependent variable. The distribution of these correlations is also
quite tight over the 30-day period of the study. Hours of sunshine tends
to have a moderately high correlation with incidence of COVID-19, but
the distribution of these correlations is more spread, and in some cases
strays into negative values. A similar thing happens with humidity,
which also tends to display more day to day variation in the correlation
with the dependent variable. The percentage of older adults shows a
relatively tight distribution of day-to-day correlations, and is
negative. Population density, in contrast, tends to be negative, but is
relatively spread, and on some days, the simple correlation between
density and incidence of COVID-19 is weakly positive. Overall for the
period under examination, the pairwise correlations between these
variables and incidence are significant at \(p<0.05\), with the
exception of the three Sunshine\_hours variables.

@@

To implement the SUR approach, we must define a matrix of spatial
weights \(W\). In this case, we consider neighborhoods based on
adjacency, based on the well-known queen criterion (two provinces are
adjacent if they share a boundary or touch at a vertex). The analysis is
of the coterminous
provinces\footnote{As a check for robustness, we also tested the rook criterion (two provinces are adjacent if they share a boundary, but not if they only touch at a vertex), and included the three islands in the sample. In this case we made an allowance for adjacency between the two islands in the Autonomous Community of Canarias in the Pacific (Las Palmas and Santa Cruz de Tenerife), and assumed that Islas Baleares in the Mediterranean are adjacent to three provinces in Pais Catalans (i.e., Barcelona, Tarragona, and Castello) The results (which can be consulted in the source R markdown document) are robust to the specification of $W$.}.

@@

After estimation, we compare the goodness of fit of the three SUR
models. Figure \ref{fig:goodness-of-fit} shows the equation-level
coefficient of determination \(R^2\), one for each time period/day. As
well, the overall coefficient of determination for the system is
reported for each model \(\text{pooled}-R^2\). The general trend is
identical for the three models, with the goodness-of-fit improving over
time and plateuing around a value of \(R^2\) of 0.55. Model 1
(\emph{lag8}) performs somewhat better in the first few equations/days,
when the goodness-of-fit is relatively poor, and then again in the last
few equations/days. Model 3 (\emph{lag11w}), in contrast, does not
perform well towards the end of the study period. The most balanced
model in terms of equation-level goodness-of-fit appears to be Model 1
(\emph{lag8}), and this impression is further supported by a slightly
higher value of the \(\text{pooled}-R^2\). The analysis using a lagged
moving average of the environmental variables is generally in line with
the incubation period reported by Lauer et al.~(2020), although the
results do not support the use of a weighted average. For the remainder
of the paper, we will adopt Model 1 (\emph{lag8}) as our best model. In
the following section we discuss the results of the analysis in more
depth.
